---
output: github_document
---

<!-- edit README.rmd -->

```{r knitr options, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# ordr

**ordr** is designed to integrate ordination analysis and biplot visualization into a [**tidyverse**](https://github.com/tidyverse/tidyverse) workflow.

## motivation

_Ordination_ is a catch-all term for a variety of classical statistical (and machine learning) techniques that decompose a numeric rectangular data set into the product of two (or more) matrices. In some cases, such as principal components analysis, the decomposition is exact; in others, such as non-negative matrix factorization, it is approximate; and in others, such as multidimensional scaling and corresppondence analysis, the data is transformed before the decomposition is performed.
What analyses that use these tools share is the inspection, and often the prediction and/or visualization, of two unlike data elements on a common set of coordinates.

The matrix decomposition usually results in at least two matrices, say $U$ and $V$, one corresponding to the rows and the other to the columns of the original data matrix $X$: If $X$ has $r$ rows and $c$ columns, then $U$ and $V$ have $r$ rows and $c$ rows, respectively, and both have $d$ columns, with $d$ being the rank of the ordination.
These $d$ columns provide a set of shared coordinates for the rows and columns of $X$. If $X$ contains a measured value for each of several variables (columns) on each of several cases (rows), then the ordination provides a space in which both cases and variables can be situated. If $X$ is a frequency table for two categorical variables, then the values of the two variables can be overlaid on the same $d$ shared coordinates. Such a scatterplot visualization of values of two unlike types is called a _biplot_, and it is the key visualization method made possible by ordination.

An extensive range of ordination techniques are already implemented in R, from classical multidimensional scaling (`stats::cmdscale()`) and principal components analysis (`stats::prcomp()` and `stats::princomp()`) in the **stats** package distributed with base R, across widely-used implementations of linear discriminant analysis (`MASS::lda()`) and correspondence analysis (`ca::ca()`) in general-use statistical packages, to highly specialized packages that implement cutting-edge techniques or tailor conventional techniques to difficult settings. These implementations come with their own conventions, tailored to the research communities that produced them, and it would be unweildy and probably even unhelpful to try to consolidate them.

Rather, **ordr** provides a streamlined process by which the outputs of these methods---in particular, the matrix factors into which the original data are approximately decomposed and the artificial coordinates they share---can be inspected, annotated, tidied, and visualized. On this last point the package may be especially helpful, since most biplot implementations provide limited customizability. The package is designed to follow the syntactic and grammatical conventions of the **tidyverse**, so that users familiar with a tidy workflow can more easily and quickly integrate ordination models into their practice.

[^svd]: Methods for `eigen()` and `svd()` decompositions are included in **ordr** for illustration. In practice, the interpretation of their output depends on the theoretical framework in which they are used.
[^lm-kmeans]: Regression and clustering models, like classical [linear regression](http://www.multivariatestatistics.org/chapter2.html) and [_k_-means](http://joelcadwell.blogspot.com/2015/08/matrix-factorization-comes-in-many.html), can also be understood as matrix decomposition approximations and visualized in biplots, on shared coordinates that are pre-defined rather than artificial. Methods for `stats::lm()` and `stats::kmeans()`, for example, are implemented for the sake of novelty and instruction, but are not widely used in practice.

## usage

### installation

**ordr** remains under development and is not scheduled for a CRAN release. For now, it can be installed using [**devtools**](https://github.com/r-lib/devtools):

```{r install}
devtools::install_github("corybrunson/ordr")
```

### example

```{r eval = TRUE, echo = FALSE, message = FALSE}
devtools::load_all()
```

Here is an example workflow that begins with a conventional PCA, proceeds through several methods provided for `tbl_ord` objects, and produces a biplot as a `ggplot` object:

```{r example}
# perform principal components analysis
(pca <- prcomp(USPersonalExpenditure, scale = TRUE))
# wrap the model as a `tbl_ord` object
(pca_ord <- as_tbl_ord(pca))
# augment the cases and variables with metadata
(pca_ord <- augment(pca_ord))
# annotate the cases or variables
(pca_ord <- mutate_v(pca_ord, year = as.integer(.name)))
# confer inertia equally between cases and variables
(pca_ord <- confer_inertia(pca_ord, c(.5, .5)))
# render a biplot, by default on the first two shared coordinates for `x` and `y`
ggbiplot(pca_ord, aes(label = .name)) +
  theme_bw() +
  geom_u_point() +
  geom_u_text_repel() +
  geom_v_vector(aes(color = year)) +
  geom_v_text_radiate() +
  ggtitle(
    "PCA of U.S. Personal Expenditure data, 1940-1960",
    "Symmetric biplot"
  )
```

## acknowledgments

### contribute

Any feedback on the package is very welcome! If you encounter confusion or errors, please create an issue with an example.

This early version only handles a few types of biplot data, a few means of transforming and annotating them, and a few plot layers. My hope is that prospective contributors can see how to expand (and improve!) on the implementations thus far.

Contributing guidelines will be posted soon. In the meantime, please respect the [**tidyverse** contributing guidelines](https://www.tidyverse.org/articles/2017/08/contributing/) and be supportive!

### inspiration

This package was originally inspired by the **ggbiplot** extention developed by [vqv](https://github.com/vqv/ggbiplot), [richardjtelford](https://github.com/richardjtelford/ggbiplot), and [GegnzaV](https://github.com/GegznaV/ggbiplot), among others. So far as i know, it first brought biplots into the **tidyverse** framework.
The motivation to unify a variety of ordination methods came from [several books and articles](https://www.barcelonagse.eu/research/publications/all?author=Michael%20Greenacre) by Michael Greenacre.
Thomas Lin Pedersen's [**tidygraph**](https://github.com/thomasp85/tidygraph) sequel to **ggraph** finally induced the shift from simply generating scatterplots to upstream handling and manipulating ordination data.

### resources

- Greenacre MJ (2010) _Biplots in Practice_. \url{http://www.multivariatestatistics.org/biplots.html}
- Palmer M ()
- Podani J (2000) _Introduction to the Exploration of Multivariate Biological Data_. \url{http://ramet.elte.hu/~podani/books.html}
- ttnphns (2015) _Answer to_ "PCA and Correspondence analysis in their relation to Biplot", CrossValidated. \url{https://stats.stackexchange.com/a/141755/68743}
