---
output: github_document
---

<!-- edit README.rmd -->

```{r knitr options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
devtools::load_all()
```

[![Travis](https://travis-ci.org/corybrunson/ordr.svg?branch=main)](https://travis-ci.org/corybrunson/ordr)

# ordr

**ordr** integrates ordination analysis and biplot visualization into [**tidyverse**](https://github.com/tidyverse/tidyverse) workflows.

## motivation

### ordination and biplots

_Ordination_ is a catch-all term for a variety of statistical techniques that introduce an artificial coordinate system for a data set in such a way that a few coordinates capture a large amount of the data structure [^ordination].
The branch of mathematical statistics called [geometric data analysis](https://www.springer.com/us/book/9781402022357) (GDA) provides the theoretical basis for (most of) these techniques.
Ordination overlaps with regression and with dimension reduction, which can be [contrasted to clustering and classification](https://towardsdatascience.com/supervised-vs-unsupervised-learning-14f68e32ea8d) in that they assign continuous rather than discrete values to data elements [^discrete].

[^ordination]: The term _ordination_ is most prevalent among ecologists; to my knowledge, no catch-all term is in common use outside ecology.
[^discrete]: This is not a hard rule: PCA is often used to compress data before clustering, and LDA uses dimension reduction to perform classification tasks.

Most ordination techniques decompose a numeric rectangular data set into the product of two matrices, often using singular value decomposition. The coordinates of the shared dimensions of these matrices (over which they are multiplied) are the artificial coordinates.
In some cases, such as principal components analysis, the decomposition is exact; in others, such as non-negative matrix factorization, it is approximate. Some techniques, such as correspondence analysis, transform the data before decomposition.
Ordination techniques may be supervised, like linear discriminant analysis, or unsupervised, like multidimensional scaling.

Analysis pipelines that use these techniques may use the artificial coordinates directly, in place of natural coordinates, to arrange and compare data elements or to predict responses. This is possible because both the rows and the columns of the original table can be located, or positioned, along these shared coordinates.
The number of artificial coordinates used in an application, such as regression or visualization, is called the _rank_ of the ordination [^lm-kmeans].
A common application is the _biplot_, which positions the rows and columns of the original table in a scatterplot in 1, 2, or 3 artificial coordinates, usually those that explain the most variation in the data.

[^lm-kmeans]: Regression and clustering models, like classical [linear regression](http://www.multivariatestatistics.org/chapter2.html) and [_k_-means](http://joelcadwell.blogspot.com/2015/08/matrix-factorization-comes-in-many.html), can also be understood as matrix decomposition approximations and even visualized in biplots. Their shared coordinates, which are pre-defined rather than artificial, are the predictor coefficients and the cluster assignments, respectively. Methods for `stats::lm()` and `stats::kmeans()`, for example, are implemented for the sake of novelty and instruction, but are not widely used in practice.

### implementations in R

An extensive range of ordination techniques are implemented in R, from classical multidimensional scaling (`stats::cmdscale()`) and principal components analysis (`stats::prcomp()` and `stats::princomp()`) in the **stats** package distributed with base R, across widely-used implementations of linear discriminant analysis (`MASS::lda()`) and correspondence analysis (`ca::ca()`) in general-use statistical packages, to highly specialized packages that implement cutting-edge techniques or adapt conventional techniques to challenging settings. These implementations come with their own conventions, tailored to the research communities that produced them, and it would be impractical (and probably unhelpful) to try to consolidate them.

Instead, **ordr** provides a streamlined process by which the models output by these methods&mdash;in particular, the matrix factors into which the original data are approximately decomposed and the artificial coordinates they share&mdash;can be inspected, annotated, tabulated, summarized, and visualized. On this last point, most biplot implementations in R provide limited customizability. **ordr** adopts the grammar of graphics paradigm from **ggplot2** to modularize and standardize biplot elements [^virtue]. Overall, the package is designed to follow the broader syntactic conventions of the **tidyverse**, so that users familiar with a this workflow can more easily and quickly integrate ordination models into practice.

[^virtue]: Biplot elments must be chosen with care, and it is useful and appropriate that many model-specific biplot methods have limited flexibility. This package adopts the trade-off articulated in [Wilkinson's _The Grammar of Graphics_](https://www.google.com/books/edition/_/iI1kcgAACAAJ) (p. 15): "This system is capable of producing some hideous graphics. There is nothing in its design to prevent its misuse. ... This system cannot produce a meaningless graphic, however."

## usage

### installation

**ordr** remains under development and is not scheduled for a CRAN release. For now, it can be installed from the (default) `main` branch using [**remotes**](https://github.com/r-lib/remotes):

```{r install, eval=FALSE}
remotes::install_github("corybrunson/ordr")
```

### example

A very common illustration of ordination in R uses Anderson's iris measurements, consisting of lengths and widths of the petals and surrounding sepals from 50 each of three species of iris:

```{r iris}
head(iris)
summary(iris)
```

An unsupervised ordination approach to these data is principal components analysis (PCA). Because a flower's sepals tend to be larger than its petals (and their lengths greater than their widths), the measurements should be scaled as well as centered before the PCA:

```{r pca}
(iris_pca <- prcomp(iris[, 1:4], scale. = TRUE))
```

**ordr** provides a new class, 'tbl_ord', that wraps ordination objects like those of 'prcomp' without directly modifying them. The original model can be recovered with `un_tbl_ord()`.

```{r tbl_ord}
(iris_pca <- as_tbl_ord(iris_pca))
```

The `print()` method for 'tbl_ord' is based on that of tibbles. Like that for **tidygraph**'s 'tbl_graph', it prints two tibbles, one for each matrix factor (in PCA, the case scores and the variable loadings of the original data matrix).
The header reminds us of the dimensions of the matrix factors and how the inertia is distributed (in PCA, with the scores in principal coordinates and the loadings in standard coordinates).

Analogous to `broom::augment()`, this tbl_ord-specific function preserves the class but augments the row and column tibbles with any metadata or diagnostics contained in the model object. Vertical bars separate the coordinate matrices from other columns.

```{r augment}
(iris_pca <- augment_ord(iris_pca))
```

Additional row- and column-level variables can be augmented and manipulated using a handful of **dplyr**-like verbs, each specific to the matrix factor being affected (`rows` or `cols`):

```{r mutate}
(iris_pca <- mutate_rows(iris_pca, Species = iris$Species))
```

Because the above steps are so common when fitting ordination models to data frames, a convenience function is available to expedite them:

```{r shortcut}
ordinate(iris, cols = 1:4, model = ~ prcomp(., scale. = TRUE))
```

Following `broom::tidy()`, this tidier for tbl_ords returns a tibble with one row per artificial coordinate (in this case, principal coordinate):[^tidy-conflict]

[^tidy-conflict]: Note that `ordr::tidy()` takes precedence for 'tbl_ord' objects because the 'tbl_ord' class precedes the underlying model class.

```{r model objects}
tidy(iris_pca)
```

The `.inertia` and `.prop_var` fields are calculated from the singular values or eigenvalues contained in the ordination object. This means that `tidy()` prepares any ordination object derived from such a decomposition for a scree plot using `ggplot2::ggplot()`:

```{r scree plot}
ggplot(tidy(iris_pca), aes(x = .name, y = .prop_var)) +
  geom_col() +
  labs(x = "", y = "Proportion of inertia") +
  ggtitle("PCA of Anderson's iris measurements",
          "Distribution of inertia")
```

Following `ggplot2::fortify()`, this fortifier combines the matrix factor tibbles into a single tibble with an additional `.matrix` column:

```{r fortification}
head(fortify(iris_pca))
tail(fortify(iris_pca))
```

The `.matrix` column is used by the row- and column-specific parts of a _dual_ layered grammar of graphics that yields a biplot:

```{r biplot}
ggbiplot(iris_pca, sec.axes = "cols", scale.factor = 2) +
  geom_rows_point(aes(color = Species, shape = Species)) +
  stat_rows_ellipse(aes(color = Species), alpha = .5, level = .99) +
  geom_cols_vector() +
  geom_cols_text_radiate(aes(label = .name)) +
  expand_limits(y = c(-3.5, NA)) +
  ggtitle("PCA of Anderson's iris measurements",
          "99% confidence ellipses; variables on top & right axes")
```

Note several conveniences: The position aesthetics are assumed to be the first and second artifical coordinates, unless otherwise specified; by default, the aspect ratio is set to 1; and an optional secondary secondary can be specified for either matrix factor and marked along the top and right borders of the plotting window.[^text-radiate]

[^text-radiate]: The radiating text geom, like other parts of the package, are adapted from the **ggbiplot** package.

## acknowledgments

### contribute

Any feedback on the package is very welcome! If you encounter confusion or errors, do create an issue, with a [minimal reproducible example](https://stackoverflow.com/help/minimal-reproducible-example) if feasible. If you have requests, suggestions, or your own implementations for new features, feel free to create an issue or submit a pull request.
Methods for additional ordination classes (see the `methods-*.r` scripts in the `R` folder) are especially welcome, as are new plot layers.
Please try to follow the
[contributing guidelines](https://github.com/corybrunson/ordr/blob/main/CONTRIBUTING.md) and respect the [Code of
Conduct](https://github.com/corybrunson/ordr/blob/main/CODE_OF_CONDUCT.md).

### inspiration

This package was originally inspired by the **ggbiplot** extension developed by [Vincent Q. Vu](https://github.com/vqv/ggbiplot), [Richard J Telford](https://github.com/richardjtelford/ggbiplot), and [Vilmantas Gegzna](https://github.com/GegznaV/ggbiplot), among others. So far as i know, it first brought biplots into the **tidyverse** framework.
The motivation to unify a variety of ordination methods came from [several books and articles](https://www.barcelonagse.eu/research/publications/all?author=Michael%20Greenacre) by Michael Greenacre, in particular [_Biplots in Practice_](https://www.fbbva.es/microsite/multivariate-statistics/resources.html#biplots).
Thomas Lin Pedersen's [**tidygraph**](https://github.com/thomasp85/tidygraph) prequel to **ggraph** finally induced the shift from the downstream generation of scatterplots to the upstream handling and manipulating of ordination models.
Additional design elements and features have been informed by the monograph [_Biplots_](https://www.routledge.com/Biplots/Gower-Hand/p/book/9780412716300) and the textbook [_Understanding Biplots_](https://www.wiley.com/en-us/Understanding+Biplots-p-9780470012550) by John C. Gower, David J. Hand, Sugnet Gardner Lubbe, and Niel J. Le Roux.

### notes

