---
title: "constructing biplot layers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{constructing biplot layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, fig.height = 7, fig.align = "center"
)
```

**ordr** originally provided custom constructors like `stat_rows_ellipse()` and `geom_cols_vector()` as the default and intended tools for adding layers to a biplot [^pre-build].
Version 0.3.0 introduces and/or documents several helper functions that instead modify either the fortified `data.frame` or the stat `ggproto`, which otherwise are used as in any `ggplot()` call.
This vignette illustrates the alternative syntaces these functions enable, which may be recommended in the next version.

[^pre-build]: The source code for these is generated before the package is built by accessing and adapting the corresponding constructors from **ggplot2** and **ggrepel** as well as several original to **ordr** (most now moved to **gggda**).

```{r setup}
library(ordr)
```


## canonical correlation analysis example

The `gilgais` data set contains measurements from regularly-spaced soil samples in New South Wales.
Hydrogen ion exponent pH (`pH`), electrical conductivity in mS/cm (`e`), and chloride content in ppm (`c`) were recorded at depths of 0--10 cm (`00`), 30--40 cm (`30`), and 80--90 cm (`80`) in soil--water extracts, yielding 9 values from each sample.
No additional metadata are provided.

```{r data}
data(gilgais, package = "MASS")
head(gilgais)
```

While important interplay is expected among these properties, none are causally determined by the others.
Ordination, rather than regression, is then a suitable approach to understanding the relationships between them.
We perform a canonical correlation analysis (CCA) of pH levels versus conductivity:

```{r cancor_ord}
gilgais_ph <- subset(gilgais, select = pH00:pH80)
gilgais_ec <- subset(gilgais, select = e00:e80)
# use `cancor_ord()` because `cancor()` returns a classless list.
gilgais_cca <- 
  ordinate(x = gilgais_ph, y = gilgais_ec, cancor_ord, scores = TRUE)
```

This technique obtains a sequence of canonical axes from each variable set that are maximally correlated.
In addition to the canonical coefficients of the manifest variables, the call above computes _canonical scores_ (coordinates in the canonical space) for the `r nrow(gilgais)` samples and the _structure correlations_ (between the manifest variables and the canonical axes).
These are stored in the `tbl_ord` object as supplementary elements:

```{r print}
print(gilgais_cca, n = 7)
```

A scree plot visualizes the quality of the fit.
We use the squares of the canonical correlations, which can be interpreted as inertia and decomposed like variance, rather than the correlations themselves:

```{r scree, fig.height=5}
tidy(gilgais_cca) %>%
  print() %>%
  ggplot(aes(x = name, y = inertia)) +
  geom_col() +
  labs(x = NULL)
```

A negligible share of inertia remains to be accounted for by the third canonical axis, so a 2-dimensional biplot will include nearly all the geometry of the variables.


## biplots

We begin with two biplots used in CCA, rendered using the row and column data separately in **ggplot2**.
The remaining sections will reproduce these biplots using different sets of tools provided by **ordr**.

```{r factor-data}
coef_ph <-
  subset(fortify(gilgais_cca), .matrix == "rows" & .element == "active")
score_ph <-
  subset(fortify(gilgais_cca), .matrix == "rows" & .element == "score")
struct_ph <-
  subset(fortify(gilgais_cca), .matrix == "rows" & .element == "structure")
coef_ec <-
  subset(fortify(gilgais_cca), .matrix == "cols" & .element == "active")
score_ec <-
  subset(fortify(gilgais_cca), .matrix == "cols" & .element == "score")
struct_ec <-
  subset(fortify(gilgais_cca), .matrix == "cols" & .element == "structure")
gilgais_inertia <- get_inertia(gilgais_cca)
```

In the appendix of _Theory and applications of correspondence analysis_ (1984), Michael Greenacre proposed a biplot of canonical scores (carrying the inertia) overlaid upon standardized canonical coefficients.
In this biplot, the projections of a sample marker onto a variable arrow _from the same matrix factor_ approximates the measurement of the variable at the corresponding sample.
Both matrix factors are handled the same way; no inertia is conferred.
The alignment of the two point clouds reflects the quality of the CCA.

```{r Greenacre-biplot}
ggplot() +
  coord_equal() +
  theme_classic() +
  geom_point(data = score_ph, aes(CanCor1, CanCor2),
             size = 1, alpha = .25, shape = "circle open") +
  geom_point(data = score_ec, aes(CanCor1, CanCor2),
             size = 1, alpha = .25, shape = "cross") +
  stat_ellipse(data = score_ph, aes(CanCor1, CanCor2)) +
  stat_ellipse(data = score_ec, aes(CanCor1, CanCor2),
               linetype = "dotted") +
  geom_vector(data = coef_ph, aes(CanCor1, CanCor2),
              linewidth = .75) +
  geom_vector(data = coef_ec, aes(100 * CanCor1, 100 * CanCor2),
              linewidth = .75, linetype = "dotted") +
  geom_text(data = coef_ph,
            aes(CanCor1, CanCor2, label = name),
            hjust = "outward", vjust = "outward") +
  geom_text(data = coef_ec, aes(100 * CanCor1, 100 * CanCor2, label = name),
            hjust = "outward", vjust = "outward") +
  scale_x_continuous(sec.axis = ~ . / 100) +
  scale_y_continuous(sec.axis = ~ . / 100)
```

In "Interpreting canonical correlation analysis through biplots of structure correlations and weights" (1990), ter Braak proposed a biplot of the structure coefficients (for both sets of manifest variables) with respect to one set of canonical axes---in our case, say, pH.
These then represent _intraset correlations_ for the pH measurements and _interset correlations_ for the conductivity measurements; as a result, the biplot fits within a unit sphere.
The sample correlation between a pH measurement and a conductivity measurement is then approximated by the inner product of the corresponding variable arrows.

```{r ter-Braak-biplot}
# confer inertia onto left variable set (pH)
struct_ec %>%
  subset(select = CanCor1:CanCor3) %>%
  sweep(2L, sqrt(gilgais_inertia), "*") %>%
  as.data.frame() %>%
  cbind(subset(struct_ec, select = -c(CanCor1:CanCor3))) ->
  struct_ec_inertia
ggplot() +
  coord_equal() +
  theme_classic() +
  # FIXME: The unit circle doesn't appear in the plot.
  geom_unit_circle() +
  scale_x_continuous(limits = c(-1, 1)) +
  scale_y_continuous(limits = c(-1, 1)) +
  geom_vector(data = struct_ph, aes(CanCor1, CanCor2), arrow = NULL) +
  geom_vector(data = struct_ec_inertia, aes(CanCor1, CanCor2),
              linetype = "dotted") +
  geom_text(data = struct_ph, aes(CanCor1, CanCor2, label = name),
            hjust = "outward", vjust = "outward") +
  geom_text(data = struct_ec_inertia, aes(CanCor1, CanCor2, label = name),
            hjust = "outward", vjust = "outward")
```

These plots serve different ends and we won't interpret them here. The remaining sections will reproduce these plots using `ggbiplot()` and 4 mutually exclusive syntactic conventions provided by **ordr**.


## custom constructors

The first convention was the first to be developed and is perhaps the easiest to use.

```{r Greenacre-biplot-constructors}
ggbiplot(gilgais_cca, sec.axes = "cols", scale.factor = 100) +
  theme_classic() +
  geom_rows_point(
    elements = "score",
    size = 1, alpha = .25, shape = "circle open"
  ) +
  geom_cols_point(
    aes(CanCor1 / 100, CanCor2 / 100), elements = "score",
    size = 1, alpha = .25, shape = "cross"
  ) +
  stat_rows_ellipse(elements = "score") +
  stat_cols_ellipse(
    aes(CanCor1 / 100, CanCor2 / 100), elements = "score",
    linetype = "dotted"
  ) +
  geom_rows_vector(linewidth = .75) +
  geom_cols_vector(linewidth = .75, linetype = "dotted") +
  geom_rows_text(aes(label = name), hjust = "outward", vjust = "outward") +
  geom_cols_text(aes(label = name), hjust = "outward", vjust = "outward")
```

The row and column constructors extract the associated data from the `tbl_ord` ordination  object but otherwise behave like standard constructors, except for the additional arguments `elements` and `subset`.
Note that secondary axes automatically apply to all elements from the appropriate matrix factor, even when this behavior is not desired, as with the scores in this plot.

```{r ter-Braak-biplot-constructors}
gilgais_cca %>%
  confer_inertia(c(0, 1)) %>%
  ggbiplot() +
  theme_classic() +
  geom_unit_circle() +
  scale_x_continuous(limits = c(-1, 1)) +
  scale_y_continuous(limits = c(-1, 1)) +
  geom_rows_vector(elements = "structure", arrow = NULL) +
  geom_cols_vector(elements = "structure", linetype = "dotted") +
  geom_rows_text(
    elements = "structure", aes(label = name),
    hjust = "outward", vjust = "outward"
  ) +
  geom_cols_text(
    elements = "structure", aes(label = name),
    hjust = "outward", vjust = "outward"
  )
```

This convention also relies on the greatest amount of package infrastructure.
Not only are new constructors and ggprotos, for example `geom_rows_vector()` and `GeomRowsVector`, written and exported to the user; the source code for them is generated by a pre-build script that can be sensitive to differences in style between extensions and even between versions of **ggplot2**.
