---
title: "constructing biplot layers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{constructing biplot layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, fig.height = 7, fig.align = "center"
)
```

**ordr** originally provided custom constructors like `stat_rows_ellipse()` and `geom_cols_vector()` as the default and intended tools for adding layers to a biplot [^pre-build].
Version 0.3.0 introduces and/or documents several helper functions that instead modify either the fortified `data.frame` or the stat `ggproto`, which otherwise are used as in any `ggplot()` call.
This vignette illustrates the alternative syntaces these functions enable, which may be recommended in the next version.

[^pre-build]: The source code for these is generated before the package is built by accessing and adapting the corresponding constructors from **ggplot2** and **ggrepel** as well as several original to **ordr** (most now moved to **gggda**).

```{r setup}
library(tidyr)
library(ordr)
theme_set(theme_classic())
```


## canonical correlation analysis example

The `gilgais` data set contains measurements from regularly-spaced soil samples in New South Wales.
Hydrogen ion exponent pH (`pH`), electrical conductivity in mS/cm (`e`), and chloride content in ppm (`c`) were recorded at depths of 0--10 cm (`00`), 30--40 cm (`30`), and 80--90 cm (`80`) in soil--water extracts, yielding 9 values from each sample.
No additional metadata are provided.

```{r data}
data(gilgais, package = "MASS")
head(gilgais)
```

While important interplay is expected among these properties, none are causally determined by the others.
Ordination, rather than regression, is then a suitable approach to understanding the relationships between them.

```{r histograms, fig.height=5}
gilgais %>%
  subset(select = c(pH00:pH80, e00:e80)) %>%
  pivot_longer(cols = seq(6), names_to = "measurement", values_to = "value") %>%
  ggplot(aes(x = value)) +
  facet_wrap(facets = vars(measurement), nrow = 2, ncol = 3, scales = "free") +
  geom_histogram() +
  labs(x = NULL, y = NULL)
```

While the distributions of pH are roughly symmetric, those of conductivity are highly right-skewed.
Ordination techniques work best in the presence of gaussian variance, so we log-transform the `e` columns before analysis.
We also select a single row on which to illustrate some optional biplot features.

```{r log-transform}
gilgais_data <- gilgais %>%
  transform(el00 = log(e00), el30 = log(e30), el80 = log(e80)) %>%
  subset(select = c(pH00:pH80, el00:el80))
head(gilgais_data)
# sample for visualization examples
i <- 4L
```

We perform a canonical correlation analysis (CCA) of pH levels versus conductivity:

```{r cancor_ord}
gilgais_ph <- subset(gilgais_data, select = pH00:pH80)
gilgais_ec <- subset(gilgais_data, select = el00:el80)
# use `cancor_ord()` because `cancor()` returns a classless list.
gilgais_cca <- 
  ordinate(x = gilgais_ph, y = gilgais_ec, cancor_ord, scores = TRUE)
```

This technique obtains a sequence of canonical axes from each variable set that are maximally correlated.
In addition to the canonical coefficients of the manifest variables, the call above computes _canonical scores_ (coordinates in the canonical space) for the `r nrow(gilgais_data)` samples and the _structure correlations_ (between the manifest variables and the canonical axes).
These are stored in the `tbl_ord` object as supplementary elements:

```{r print}
print(gilgais_cca, n = 7)
```

A scree plot visualizes the quality of the fit.
We use the squares of the canonical correlations, which can be interpreted as inertia and decomposed like variance, rather than the correlations themselves:

```{r scree, fig.height=5}
tidy(gilgais_cca) %>%
  print() %>%
  ggplot(aes(x = name, y = inertia)) +
  geom_col() +
  labs(x = NULL)
```

A negligible share of inertia remains to be accounted for by the third canonical axis, so a 2-dimensional biplot will include nearly all the geometry of the variables.


## biplots

We begin with two biplots used in CCA, rendered in **ggplot2** using the row and column data prepared in the code chunk below.
The remaining sections will reproduce these biplots using different sets of tools provided by **ordr**.

```{r factor-data}
# variable loadings onto canonical axes
coef_ph <-
  subset(fortify(gilgais_cca), .matrix == "rows" & .element == "active")
coef_ec <-
  subset(fortify(gilgais_cca), .matrix == "cols" & .element == "active")
# case scores onto canonical axes
score_ph <-
  subset(fortify(gilgais_cca), .matrix == "rows" & .element == "score")
score_ec <-
  subset(fortify(gilgais_cca), .matrix == "cols" & .element == "score")
# structure coefficients (correlations of variables with canonical axes)
struct_ph <-
  subset(fortify(gilgais_cca), .matrix == "rows" & .element == "structure")
struct_ec <-
  subset(fortify(gilgais_cca), .matrix == "cols" & .element == "structure")
# extract inertia and confer onto all elements
gilgais_inertia <- get_inertia(gilgais_cca)
confer_var <- function(x) {
  x %>%
    subset(select = CanCor1:CanCor3) %>%
    sweep(2L, sqrt(gilgais_inertia), "*") %>%
    as.data.frame() %>%
    cbind(subset(x, select = -c(CanCor1:CanCor3)))
}
coef_ph_inertia <- confer_var(coef_ph)
score_ph_inertia <- confer_var(score_ph)
struct_ph_inertia <- confer_var(struct_ph)
coef_ec_inertia <- confer_var(coef_ec)
score_ec_inertia <- confer_var(score_ec)
struct_ec_inertia <- confer_var(struct_ec)
```

For convenience, we also pre-define two distinct arrow styles:

```{r arrows}
arrow_closed <- arrow(length = unit(.03, "native"), type = "closed")
arrow_opened <- arrow(length = unit(.03, "native"), type = "open")
```

### ter Braak CCA biplot

In "Interpreting canonical correlation analysis through biplots of structure correlations and weights" (1990), ter Braak proposed a biplot of the structure coefficients (for both sets of manifest variables) with respect to one set of canonical axes---in our case, say, pH.
These then represent _intraset correlations_ for the pH measurements and _interset correlations_ for the conductivity measurements; as a result, the biplot fits within a unit sphere.
The sample correlation between a pH measurement and a conductivity measurement is then approximated as the inner product of the corresponding variable arrows.

```{r ter-Braak-biplot, fig.width=5.5, fig.height=5}
ggplot(mapping = aes(CanCor1, CanCor2)) +
  coord_equal() +
  geom_unit_circle() +
  scale_x_continuous(limits = c(-1, 1)) +
  scale_y_continuous(limits = c(-1, 1)) +
  geom_vector(data = struct_ph, arrow = arrow_closed) +
  geom_vector(data = struct_ec_inertia, arrow = arrow_opened,
              linetype = "dashed") +
  geom_text(data = struct_ph, aes(label = name),
            hjust = "outward", vjust = "outward") +
  geom_text(data = struct_ec_inertia, aes(label = name),
            hjust = "outward", vjust = "outward")
```

The biplot indicates that pH and conductivity are highly correlated closer to the surface but highly anticorrelated at lower depths---the latter with more confidence than the former, as the vectors lie more nearly within the plane of the first two canonical axes.
There is also a general positive correlation among conductivity measurements at different depths, in contrast to the largely uncorrelated depth-specific pH measurements.

### Greenacre CCA biplot

In the appendix of _Theory and applications of correspondence analysis_ (1984), Michael Greenacre proposed a biplot of canonical scores (carrying the inertia) overlaid upon standardized canonical coefficients.
The axes are recalibrated to undo the centering operation, so that the projection of a sample marker onto a variable arrow _from the opposing matrix factor_ approximates the measurement of the variable at the corresponding sample.
**TODO: This needs to be shown algebraically.**
Both matrix factors are handled the same way; no inertia is conferred.
The alignment of the two point clouds reflects the quality of the CCA.

```{r Greenacre-biplot, fig.height=4}
ggplot(mapping = aes(CanCor1, CanCor2)) +
  coord_equal() +
  geom_text(data = score_ph_inertia, aes(label = name),
            size = 2, alpha = 1/3) +
  geom_axis(data = coef_ec_inertia,
            aes(CanCor1 * 100, CanCor2 * 100, center = center, label = name)) +
  stat_projection(data = score_ph_inertia[i, , drop = FALSE],
                  referent = coef_ec_inertia[c(1, 3), , drop = FALSE],
                  linetype = "dotted") +
  scale_x_continuous(sec.axis = ~ . / 100) +
  scale_y_continuous(sec.axis = ~ . / 100)
```

Projections are shown from sample `r i` onto the axes for the extremal depths.
The predictions are not far from the mark, though we must exp-transform the `el` values to get their original units.

These plots serve different ends and we won't interpret them further here. The remaining sections will reproduce these plots using `ggbiplot()` and 4 mutually exclusive syntactic conventions provided by **ordr**.


## custom constructors

The first convention was the first to be developed and is perhaps the easiest to use.

```{r ter-Braak-biplot-constructors, fig.width=5.5, fig.height=5}
gilgais_cca %>%
  confer_inertia(c(0, 1)) %>%
  ggbiplot() +
  geom_unit_circle() +
  scale_x_continuous(limits = c(-1, 1)) +
  scale_y_continuous(limits = c(-1, 1)) +
  geom_rows_vector(elements = "structure", arrow = arrow_closed) +
  geom_cols_vector(elements = "structure", arrow = arrow_opened,
                   linetype = "dashed") +
  geom_rows_text(
    elements = "structure", aes(label = name),
    hjust = "outward", vjust = "outward"
  ) +
  geom_cols_text(
    elements = "structure", aes(label = name),
    hjust = "outward", vjust = "outward"
  )
```

The row and column constructors extract the associated data from the `tbl_ord` ordination object but otherwise behave like standard constructors, except for the additional arguments `elements` and `subset`.
Note that the inertia conferred onto the rows and columns applies to all row and column elements, respectively, so does not have to be bookkept as by the passage to `data` arguments above.

```{r Greenacre-biplot-constructors, fig.height=4}
gilgais_cca %>%
  confer_inertia(c(1, 1)) %>%
  ggbiplot(sec.axes = "cols", scale.factor = 100) +
  geom_rows_text(aes(label = name), elements = "score", size = 2, alpha = 1/3) +
  geom_cols_axis(aes(center = center, label = name)) +
  stat_rows_projection(
    elements = "score", subset = i, ref_subset = c(1, 3),
    linetype = "dotted"
  )
```

Note that secondary axes automatically apply to all elements from the appropriate matrix factor, even when this behavior is not desired, as with the scores in this plot.
Above, in contrast, the constructors in need of specific positional aesthetics were those we wished to respect the secondary axes.

This convention also relies on the greatest amount of package infrastructure.
Not only are new constructors and ggprotos, for example `geom_rows_vector()` and `GeomRowsVector`, written and exported to the user; the source code for them is generated by a pre-build script that can be sensitive to differences in style between extensions and even between versions of **ggplot2**.
Because not all conceivable constructors are pre-built, users are limited in what plot elements they can declare this way.

## stat specializers

Since the beginning, **ordr**'s custom constructors have been supported by the `rows_stat()` and `cols_stat()` helpers, which convert conventional stat ggprotos or their names to the names for their factor-specific counterparts.
For example, `rows_stat(StatEllipse)` and `rows_stat("ellipse")` both return `"rows_ellipse"`.
While these helpers were originally unexported and required input, they are now exported and default to their corresponding identity stats, `GeomRowsIdentity` and `GeomColsIdentity`.

```{r ter-Braak-biplot-specializers, fig.width=5.5, fig.height=5}
gilgais_cca %>%
  confer_inertia(c(0, 1)) %>%
  ggbiplot() +
  geom_unit_circle() +
  scale_x_continuous(limits = c(-1, 1)) +
  scale_y_continuous(limits = c(-1, 1)) +
  geom_vector(stat = rows_stat(), elements = "structure",
              arrow = arrow_closed) +
  geom_vector(stat = cols_stat(), elements = "structure",
              arrow = arrow_opened, linetype = "dashed") +
  geom_text(
    stat = rows_stat(), elements = "structure", aes(label = name),
    hjust = "outward", vjust = "outward"
  ) +
  geom_text(
    stat = cols_stat(), elements = "structure", aes(label = name),
    hjust = "outward", vjust = "outward"
  )
```

The syntax is cumbersome but natural when the underlying stat is the identity.
While less efficient than using the custom constructors, the lines above may be easier to understand for someone familiar with **ggplot2** but just learning **ordr**.

```{r Greenacre-biplot-specializers, fig.height=4}
gilgais_cca %>%
  confer_inertia(c(1, 1)) %>%
  ggbiplot(sec.axes = "cols", scale.factor = 100) +
  geom_text(
    aes(label = name),
    stat = rows_stat(), elements = "score", size = 2, alpha = 1/3
  ) +
  geom_axis(stat = cols_stat(), aes(center = center, label = name)) +
  geom_segment(
    stat = rows_stat("projection"),
    elements = "score", subset = i, ref_subset = c(1, 3),
    linetype = "dotted"
  )
```

A more significant deficiency of these helpers is that they only apply to stats, which means that each layer must be declared using a geom constructor (or else a `layer()` call).
Even i had to check the source code for `stat_projection()`, to remind myself that it couples by default with the segment geom, in order to declare the projection layers above.[^convenience-constructors]

[^convenience-constructors]: Many "in-grammar" **ggplot2** extensions provide convenience constructors like `geom_projection()`, modified from existing constructors like `geom_segment()` only to default to a different stat. These can be especially helpful to the likely majority of **ggplot2** users who don't fully understand the underlying grammar. However, they also implicitly conflate stats and geoms (as equivalent to "layers" rather than as components of them) and foreclose the use of novel stat--geom pairings. (It's also remarkably rare, in my experience, to see convenience `stat_*()` constructors for new geoms.) For these reasons, i avoid this practice.

These helpers could also be used to declare additional plot layers using `layer()` directly. This is, i think, an underappreciated option, but one that does not require a separate demonstration here.
While these helpers allow for a syntax that may be more intuitive for new users, they do not provide any additional functionality, and may inhibit the use of valuable layers by users not habituated into developer habits like inspecting source code (or, i guess, reading the manual).

## ggproto generators

The ggproto constructors `make_rows_stat()` and `make_cols_stat()` go a significant step further:
In case no row- or column-specific ggproto is found, they can be used to create one.
The constructors follow a standard sequence of steps to incorporate ordination aesthetics and, if necessary, referential parameters into a new child ggproto, which also filters the input data by `.matrix`.
They can be used in the interactive environment à la `StatRowsConnect <- make_rows_stat(StatConnect)`, but here we use them to generate the necessary ggprotos "on the fly" rather than rely on their having been written and exported in **ordr**.

```{r ter-Braak-biplot-generators, fig.width=5.5, fig.height=5}
gilgais_cca %>%
  confer_inertia(c(0, 1)) %>%
  ggbiplot() +
  geom_unit_circle() +
  scale_x_continuous(limits = c(-1, 1)) +
  scale_y_continuous(limits = c(-1, 1)) +
  geom_vector(stat = make_rows_stat(), elements = "structure",
              arrow = arrow_closed) +
  geom_vector(stat = make_cols_stat(), elements = "structure",
              arrow = arrow_opened, linetype = "dashed") +
  geom_text(
    stat = make_rows_stat(), elements = "structure", aes(label = name),
    hjust = "outward", vjust = "outward"
  ) +
  geom_text(
    stat = make_cols_stat(), elements = "structure", aes(label = name),
    hjust = "outward", vjust = "outward"
  )
```

When not needed, these constructors should be avoided; they take up more space than the specializers while doing unnecessary extra work.

```{r Greenacre-biplot-generators, fig.height=4}
gilgais_cca %>%
  confer_inertia(c(1, 1)) %>%
  ggbiplot(sec.axes = "cols", scale.factor = 100) +
  geom_text(
    aes(label = name),
    stat = make_rows_stat(), elements = "score", size = 2, alpha = 1/3
  ) +
  geom_axis(stat = make_cols_stat(), aes(center = center, label = name)) +
  geom_segment(
    stat = make_rows_stat("projection"),
    elements = "score", subset = i, ref_subset = c(1, 3),
    linetype = "dotted"
  )
```

While not necessary, the constructor demonstrates its compatibiilty with reference layers like that of projection: The `$setup_params` method is redefined to accommodate reference parameters.

In principle, any statistical transformation ggproto should be adaptable in this way, provided it includes numeric `x` and `y` coordinates as output.
This opens enormous potential to **ordr** biplots that could not practically be implemented or maintained through custom constructors.
By also enabling users to assign ggproto source code to suitable names like `StatRowsConnect`, these constructors may provide value even beyond convenient biplot layering.

## fortified data subsetters

The use of `fortify()` by `ggplot()` is often perfunctory.
It converts certain model objects to data frames but leaves data frames as they are, so has no effect on the vast majority of `ggplot()` calls.
In contrast, `ggbiplot()` is designed to rely almost always on the `tbl_ord` method of `fortify()`.
This method row-binds the matrix factors of a wrapped ordination model to obtain a data frame, the source factor (`"rows"` or `"cols"`) marked by a new `.matrix` column:

```{r fortify}
gilgais_df <- fortify(gilgais_cca)
head(gilgais_df)
tail(gilgais_df)
```

The custom constructors above use this `.matrix` column to subset the fortified data before applying statistical transformation and geometric construction.
An alternative option is to do this manually via the standard `data` argument:
Instead of a data frame, `data` can also receive a function, which is then applied to the data frame inherited by the layer.
The subsetting helpers `rows_data()` and `cols_data()` are little function factories; they each return a function that subsets an incoming data frame to its `"rows"` or `"cols"` rows.
We pass these factory calls rather than functions so that we can include `subset` and `elements` arguments to be applied to `data`.

```{r ter-Braak-biplot-subsetters, fig.width=5.5, fig.height=5}
gilgais_cca %>%
  confer_inertia(c(0, 1)) %>%
  ggbiplot() +
  geom_unit_circle() +
  scale_x_continuous(limits = c(-1, 1)) +
  scale_y_continuous(limits = c(-1, 1)) +
  geom_vector(data = rows_data(elements = "structure"),
              arrow = arrow_closed) +
  geom_vector(data = cols_data(elements = "structure"),
              arrow = arrow_opened, linetype = "dashed") +
  geom_text(
    data = rows_data(elements = "structure"), aes(label = name),
    hjust = "outward", vjust = "outward"
  ) +
  geom_text(
    data = cols_data(elements = "structure"), aes(label = name),
    hjust = "outward", vjust = "outward"
  )
```

This syntax doesn't meaningfully shorten lines of code.
It also demands that the `data` argument be named in the call, except on rare occasions that aesthetic mappings are passed first.

```{r Greenacre-biplot-subsetters, fig.height=4}
gilgais_cca %>%
  confer_inertia(c(1, 1)) %>%
  ggbiplot(sec.axes = "cols", scale.factor = 100) +
  geom_text(
    aes(label = name),
    data = rows_data(elements = "score"), size = 2, alpha = 1/3
  ) +
  geom_axis(data = cols_data(), aes(center = center, label = name)) +
  stat_projection(
    data = rows_data(elements = "score", subset = i),
    referent = cols_data(subset = c(1, 3)),
    linetype = "dotted"
  )
```

This option also introduces some risky flexibility:
Whereas `stat_rows_projection()` automatically takes the opposing `"cols"` matrix factor as the referent data (and vice-versa), here the user must specify `referent = cols_data()` in order to obtain a correct biplot.

Despite their problems, these subsetters allow users to add any sufficiently well-behaved statistical transformation ggproto to a biplot layer.
Like the specialized ggproto constructors, this eliminates the practical barriers to biplots using stats and geoms from other extensions or even those custom-built by users.
It also eliminates the burden of creating those specialized ggprotos at definition time.
The trade-off is that the user must assume responsibility for some biplot-specific behavior.
